\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-04 trunk/europa-basics.tex
updates

The most common planning formulations use a propositional representation, where the state of the world is represented by a set of propositions (statements that can be true or false), and operators change the truth values of these propositions (TODO ref?). Although these formulations are powerful and have allowed researchers to develop numerous contributions in automated planning, there are many classes of problem domains that are difficult it is difficult to represent using them.

In particular, it is hard to represent time, resources, mutual exclusion and concurrency using propositions (TODO: ref?).  It is straightforward to represent and reason about all of those elements using a CSP representation, as a result there has been some work on doing automated planning while taking advantage of CSP  (TODO: ref).  Traditionally, the entire planning problem is translated into a CSP and then solved using traditional CSP methods, this leads to a formulation where action choices and relationships are represented as variables and/or constraints. However, this has at least two major drawbacks:
\begin{enumerate}
	\item Given that action choices and relationships (rules) are expresed through variables, the domain descriptions that result from this approach are not intuitive and therefore difficult to understand and debug
	\item If the structure of a planning model (actions, conditions, effects, dependencies at the action level, etc) is not explicitly maintained by the CSP planner, the search algorithms are deprived of critical information to make better decisions. If that structure is maintained somehow (for instance, by internally marking variables that represent action choice and relationships between actions), it is still hard to write search algorithms as any planning-specific insight has to be translated into the CSP representation of variables and constraints  (TODO: example?)
\end{enumerate}

Constraint-Based Attribute and Interval planning (CAIP) (TODO:ref) is intended to close that gap: on the one hand, it takes advantage of a CSP representation, but it also uses attributes and intervals to maintain an explicit representation of the elements relevant to planning, so that it is easier to write algorithms that search for an reason abut plans. The main elements of CAIP are:

\begin{enumerate}
	\item Intervals:
	\item Attributes:
	\item Domain Constraints an Configuration rules:
\end{enumerate}

TODO: Explain how planning problem instances and their solutions are represented.

>>>>

The most common planning formulations use a propositional
representation, where the state of the world is represented by a set
of propositions (statements that can be true or false), and operators
change the truth values of these propositions \cite{gen87}. Although
these formulations are powerful and have allowed researchers to
develop numerous contributions in automated planning, there are many
classes of problem domains that are difficult to represent using this
formalism. In particular, it is hard to represent time, resources,
mutual exclusion and concurrency using propositions \comment{citation
  needed}. While CSP representations have traditionally had an edge in
formulating and representing planning problems, 
% It is straightforward to represent and reason about all of
% those elements using a CSP representation, as a result there has been
% some work on doing automated planning while taking advantage of CSP
% (TODO: ref). Traditionally, the entire planning problem is translated
% into a CSP and then solved using traditional CSP methods, this
leading to a formulation where action choices and relationships are
represented as variables and/or constraints, there are at least two
major drawbacks:

\begin{enumerate}

\item Given that action choices and relationships (rules) are expressed
  through variables, the domain descriptions that result from this
  approach are not intuitive and therefore difficult to understand and
  debug

\item If the structure of a planning model (actions, conditions,
  effects, dependencies at the action level) is not explicitly
  maintained by the CSP planner, the search algorithms are deprived of
  critical information to make better decisions. If that structure is
  maintained (for instance, by internally marking variables that
  represent action choice and relationships between actions), it is
  still hard to write search algorithms as any planning-specific
  insight has to be translated into the CSP representation of
  variables and constraints \comment{For example?}

Constraint-based Planning and specifically Constraint-Based Attribute
and Interval planning (CAIP) \cite{mus94,frank2003} is intended to
close that gap. On one hand, it takes advantage of a CSP
representation; one the other it also uses attributes and intervals to
maintain an explicit representation of the elements relevant to
planning. This makes it easier to write algorithms that search for an
reason abut plans. The main elements of CAIP are:

\begin{enumerate}
	\item Intervals: \comment{needs to be defined}
	\item Attributes:
	\item Domain Constraints an Configuration rules:
\end{enumerate}

\comment{Explain how planning problem instances and their solutions are represented}

\subsection{\eu Plan Representation}
\label{sec:europa:pr}

\eu is an implementation of the CAIP framework.

Variables

Values that need to be represented to describe the problem domain and over which we may want to specify constraints.

Objects

The things we wish to describe and refer to in a domain are considered
Objects. As in the case with object-oriented analysis and design, one
can seek out the nouns in any domain description to find likely
objects. In a Mars rover example, we might consider the satellite, the
camera, and the attitude controller (which orients the robot for
taking pictures) to be objects. Objects have state and behavior. For
example, a camera can be:

    off,
    ready, or
    taking a picture.

An attitude controller can be:

    pointing at a position, or
    turning from one position to another.

    An object is an instance of a class. In EUROPA we model using the
    abstraction of a class to speak about all instances having certain
    properties of state and behavior. In order to describe such state
    and behavior we build on the formalism of first order logic.

Temporally Qualified Predicates and Actions

A predicate defines a relation between objects and properties. In
EUROPA, we define such relations between variables whose domains are
sets of objects and sets of properties. For example, we might use a
predicate Pointing(a,p) to indicate that a rover's attitude controller
a is pointing at position p. Note that a is a variable which may have
a number of possible values in a problem with multiple
satellites. Similarly, p is a variable whose values are the set of
possible positions. In a grounded plan, single values will be
specified for each variable as we saw in the case of a solved
Constraint Satisfaction Problem.

In general, is is not sufficient to state that a predicate is true
without giving it some temporal extent over which it holds. Predicates
that are always true can be thought to hold from the beginning to the
end of time. However, in practice, the temporal extent of interest
must be defined with timepoints to represent its start and end. So we
might want to write Pointing(a,s,e,p) to indicate that the attitude
controller a is pointing at position p from time s to time e. In fact,
this pattern of using such predicates to describe both state and
behavior of objects is so prevalent in EUROPA that we have introduced
a special construct called a Token which has the built in variables to
indicate the object to which the statement principally applies and the
timepoints over which it holds. In EUROPA, all predicate instances are
Tokens.

Tokens

A Token is an instance of a predicate or an action and is defined over
a temporal extent. Every token has five built-in variables:

start: The beginning of the temporal extent over which the predicate
is defined.  end: The end of the temporal extent over which the
predicate is defined.  duration: The constraints start + duration =
end is enforced automatically.  object: made. The set of objects to
which a token might apply. In a grounded plan each Token applies to a
specific Object, reflecting the intuition that we are using Tokens to
describe some aspect of an Object (i.e. its state or behavior) in
time. However, in a partial plan, the commitment to a specific object
may not yet have been made.

state: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
variable captures the token's current state and its reachable states
through further restriction. See the Token State Model discussion
below for details.

Objects (continued)

As we have noted, Tokens describe some aspect of an Object in
time. Objects thus may have many Tokens in a plan in order to describe
their state and behavior throughout all points in time of the
plan. Within this general framework, we note a few particulars:

Static Facts: Classes without predicates lead to Objects without
Tokens. This arises where an objects state or behavior is independent
of time. For example, a domain may have a set of locations and/or
paths for which there is nothing more to say than that they exist.

Timelines: Often objects in a domain must be described by exactly one
token for every given timepoint in the plan. Such objects are so
common that we provide a special construct in EUROPA to extend these
semantics to derived classes. Any instances of a class derived from a
Timeline will induce ordering requirements among its tokens in order
to ensure no temporal overlap may occur among them. See below for
further discussion.

Resources: Metric resources, e.g. the energy of a battery or the
capacity of a cargo hold, are objects with an explicit quantitative
state in time and with a circumscribed range of changes that can occur
to impact that state i.e. produce, consume, use, change. These changes
are captured as tokens. Resources are such a common requirement for
EUROPA users that special constructs are also provided for
them. Instances of classes derived from a Resource will induce
ordering requirements on their Tokens in order to ensure that the
level of the resource remains within specified limits.

Timelines

It may be sufficient to maintain only a partial-order among
tokens. However, it is often the case that tokens represent states and
actions of a single object in the system. Such tokens are typically
mutually-exclusive. EUROPA uses a Timeline structure developed in (??)
to concisely capture system components whose behavior is described
over time in this manner.

For example, consider the tire-world domain. The tire was located in
the trunk with the predicate tireLocated(Trunk) and located on the
ground with the predicate tireLocated(Ground). Clearly, the same tire
cannot be in both places at once. A Timeline provides a simple method
for aggregating the statements about a tire such that they are
mutually exclusive.

Figure 1: A Timeline for a Tire

Figure 1 illustrates how a Timeline can be used to describe the
whereabouts of a tire. The tire is an object in the tire-world
represented as a timeline. It has predicates associated with it which
can describe its states and actions over time. The predicate names
need not contain the tire prefix; this is now implicit since the
tokens are assigned to a specific tire instance (i.e. an instance of a
Timeline). While Timelines are a useful element of the EUROPA planning
paradigm, they are not essential. A more general notion of a system
Object can be used which does not impose restrictions of
mutual-exclusion or non-zero duration. This can be important in
supporting partial-order planning.

For this example, the state of the tire is specified using 3
contiguous tokens. The end and start time-points are related by an
equality constraint. A Moving predicate has been introduced to cover
the transition from one location to another. It takes 2 location
arguments. Notice that precedence relationships exist between tokens
such that they cannot overlap but the start and end times may remain
flexible. To illustrate this, sample times are included. Assume a
total time range of interest between 0 and 1000. In addition, tokens
on Timelines have a minimum duration of 1. As a result the values
shown are the most flexible possible values for the start and end of
each token. There are a number of advantages of allowing this
flexibility. First, the basic structure of the plan can be developed
without over-committing to specific times. If it is not necessary for
the tire to be on the ground at time-step 4, then a planner should not
be forced to specify it. Such an approach permits a least-commitment
approach to planning. Second, in many domains it is simply impossible
to know in planning exactly how long an activity might take. For
example, a common activity of driving a car from one point to another
on a road can take varying amounts of time depending on traffic and
traffic lights. In such circumstances it is more practical to express
upper and lower bounds on durations which naturally lead to intervals
for start and end times. In such domains, flexibility in planning aids
robustness in execution.  Rules

In order for a plan to be valid, it must comply with all rules and
regulations pertinent to the application domain in question. Rules
govern the internal and external relationships of a token. For
example, consider a parameterized predicate describing a transition
from one location to another. Let the parameters be from and to. The
parameters are instantiated on a token as variables whose domain of
values is the set of all locations in a given problem. A rule
governing an internal relationship among token variables might
stipulate that a transition must involve a change in location. This
can be easily expressed as a constraint on the definition of a
predicate of the form: from != to. It is reasonable to further
stipulate that one must be Located somewhere before a transition can
occur, and one must end up Located somewhere when completed. This is
an example of a rule governing an external relationship among
tokens. It specifies a requirement that tokens of the predicate
Located precede and succeed tokens of the predicate Moving.

Figure 2: Internal and external relationships from rules on Moving
Figure 2 illustrates the entities and relations involved in specifying
such a rule on a Moving predicate. The token on which the rule applies
is referred to as the master. Each Located token required by the
master is referred to as a slave. All variables are indicated by name
and their domains are expressed as intervals in the case of temporal
variables and as enumerations for the remainder. Application of a rule
on a token can thus cause slave tokens, variables, and constraints to
be introduced.  Token State Model

The capability of a domain rule to cause a slave token to be created
is a key vehicle through which planning occurs. Semantically, this
rule imposes a requirement for supporting tokens to be in the plan in
order for the master to be valid. There are 2 possibilities to
consider:

    The slave is inserted as an active token in the plan. As such, rules may be activated on the slave, and it may consume resources.
    The token is merged with a matching token already in the plan. Once a slave is merged, the requirement it represents are considered satisfied. The process of merging passes on all restrictions imposed on the slave to the active token upon which it is merged. Merging a token requires finding a target active token that is compatible with the inactive token. For an active token and an inactive token to be compatible requires that they are instances of the same predicate and that no intersections between corresponding variables are empty. The effect of merging is illustrated in Figure 3.

    Figure 3: Merging an inactive token on an active token. Domain
    restrictions occur in highlighted variables of active token.

    On creation of a token, where the commitment has not been made yet
    to activate or merge the token, the token is said to be
    inactive. There is a nuance to the state model for tokens which
    relate to the mode of its creation. If the token is allocated
    explicitly by an external actor, rather than internally through
    rule firing, it may include the state rejected indicating the
    planner is permitted to reject the token. A valid plan can include
    rejected tokens. This typically arises where the token represents
    a goal that is preferable to achieve but not mandatory. This state
    is not reachable for slaves since that would imply selective
    adherence to the domain model. Figure 4 presents the state
    transition diagram for token states relevant for planning. The
    transitions are operations on a partial plan which can decide an
    outcome for an inactive token. As operations which may arise in
    search, they must be reversible during backtracking. The
    cancellation operations in each case are also shown.

Figure 4: Token States and Transitions for Planning

The state of a token is embodied with a 5th and final built-in
variable referred to as the state variable. The planner states are
values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
INACTIVE state is captured by the variable being unbound. If a value
is removed from the domain, that state will not be reachable. For
example, when a slave is created via a rule, the REJECTED value is
removed from its state variable.  Summary

This page has presented the main elements of the EUROPA planning approach. They are:

Temporally scoped predicates and actions (a.k.a. Tokens) to represent
states and actions in time. Note that tokens do not discriminate
between state and action.

Constraints to describe relationships among tokens. This provides an
expressive method of describing interactions among tokens in the
context of Temporal Planning.

Timelines as a concise abstraction to express the evolution of state
and behavior for a system component. It provides semantics of mutual
exclusion sequencing in time. Other core abstractions are available in
EUROPA for handling metric resources.

Domain rules to describe internal and external relationships on and
between tokens respectively. Rules are applied to active tokens
(master tokens) referred to as masters and typically produce inactive
tokens (slaves) which are a key vehicle through which the planning
process evolves. The rule structure in EUROPA differs from the more
restrictive commitment to preconditions and effects in classical
planning which prohibits durative actions and disembodied effects
(i.e. effects which may not occur until some temporal distance after
the end of the action).

Token States which are the basis of planning operations on a partial
plan.
